---
title: "HIV Treatment Cascade Analysis"
author: "Brittany Bowman"
date: "9/26/2019"
output: html_document
runtime: shiny
---

This file is an analysis of HIV treatment cascade data from UNAIDS from 2015-2018 related to progress towards the 90-90-90 targets.

```{r setup, include=FALSE echo = FALSE}
# load necessary packages
library(tidyverse)
library(shiny)
library(plotly)
library(scales)

# get working directory
getwd()

# import regional data sets
treat_cascade_all_global <- read_csv("data/treatment_cascade_global_all_ages.csv")
treat_cascade_m_adult_global <- read_csv("data/treatment_cascade_global_male_adults.csv")
treat_cascade_f_adult_global <- read_csv("data/treatment_cascade_global_female_adults.csv")
test_treat_cascade_all_global <- read_csv("data/test_and_treat_cascade_global_all_ages.csv")
test_treat_cascade_m_adult_global <- read_csv("data/test_and_treat_cascade_global_male_adults.csv")
test_treat_cascade_f_adult_global <- read_csv("data/test_and_treat_cascade_global_female_adults.csv")
know_status_all_global <- read_csv("data/know_status_global_all_ages.csv")
on_art_all_global <- read_csv("data/receive_art_global_all_ages.csv")
vl_suppress_all_global <- read_csv("data/vl_supression_total_hiv_global_all_ages.csv")

# import national data sets
treat_cascade_all <- read_csv("data/treatment_cascade_all_ages.csv")
treat_cascade_m_adult <- read_csv("data/treatment_cascade_male_adults.csv")
treat_cascade_f_adult <- read_csv("data/treatment_cascade_female_adults.csv")
test_treat_cascade_all <- read_csv("data/test_and_treat_cascade_all_ages.csv")
test_treat_cascade_m_adult <- read_csv("data/test_and_treat_cascade_male_adults.csv")
test_treat_cascade_f_adult <- read_csv("data/test_and_treat_cascade_female_adults.csv")
know_status_all <- read_csv("data/know_status_all_ages.csv")
on_art_all <- read_csv("data/receive_art_all_ages.csv")
vl_suppress_all <- read_csv("data/vl_supression_total_hiv_all_ages.csv")
```


```{r explore data}
# create list of global datasets
global_data_list <- list(treat_cascade_all_global, treat_cascade_m_adult_global, treat_cascade_f_adult_global, test_treat_cascade_all_global, test_treat_cascade_m_adult_global, test_treat_cascade_f_adult_global, know_status_all_global, on_art_all_global, vl_suppress_all_global)

global_list_names <- c("treat_cascade_all_global", "treat_cascade_m_adult_global", "treat_cascade_f_adult_global", "test_treat_cascade_all_global", "test_treat_cascade_m_adult_global", "test_treat_cascade_f_adult_global", "know_status_all_global", "on_art_all_global", "vl_suppress_all_global")

names(global_data_list) <- global_list_names

# create list of country datasets
country_data_list <- list(treat_cascade_all, treat_cascade_m_adult, treat_cascade_f_adult, test_treat_cascade_all, test_treat_cascade_m_adult, test_treat_cascade_f_adult, know_status_all, on_art_all, vl_suppress_all) 

country_list_names <- c("treat_cascade_all", "treat_cascade_m_adult", "treat_cascade_f_adult", "test_treat_cascade_all", "test_treat_cascade_m_adult", "test_treat_cascade_f_adult", "know_status_all", "on_art_all", "vl_suppress_all") 

names(country_data_list) <- country_list_names

# view raw datasets
map(global_data_list, glimpse)
map(country_data_list, glimpse)

# clear lists to free space
rm(global_data_list, country_data_list)
```


```{r create new composite data frames}
# create functions for cleaning test_treat datasets to combine with treat datasets
trim_test_treat_global <- function(data) {
  data %>%
    select(Region = X1, ends_with("_1"), ends_with("_2")) %>%
    rename_at(vars(ends_with("_1")), str_replace, "_1", "_hiv_on_art") %>% 
    rename_at(vars(ends_with("_2")), str_replace, "_2", "_hiv_vl_suppress") %>%
    filter(Region != "Region")
}

trim_test_treat_country <- function(data) {
  data %>%
    select(Country = X1, ends_with("_1"), ends_with("_2")) %>%
    rename_at(vars(ends_with("_1")), str_replace, "_1", "_hiv_on_art") %>% 
    rename_at(vars(ends_with("_2")), str_replace, "_2", "_hiv_vl_suppress") %>%
    filter(Country != "Country")
}

# create functions for cleaning treat datasets to combine with test_treat datasets
trim_treat_global <- function(data) {
  data %>%
    rename(Region = X1) %>%
    rename_at(vars(-Region, -ends_with("_1"), -ends_with("_2")), str_c, "_know_status") %>% 
    rename_at(vars(ends_with("_1")), str_replace, "_1", "_know_status_on_art") %>% 
    rename_at(vars(ends_with("_2")), str_replace, "_2", "_on_art_vl_suppress") %>%
    filter(Region != "Region")
}

trim_treat_country <- function(data) {
  data %>%
    rename(Country = X1) %>%
    rename_at(vars(-Country, -ends_with("_1"), -ends_with("_2")), str_c, "_know_status") %>% 
    rename_at(vars(ends_with("_1")), str_replace, "_1", "_know_status_on_art") %>% 
    rename_at(vars(ends_with("_2")), str_replace, "_2", "_on_art_vl_suppress") %>%
    filter(Country != "Country")
}

# create lists of composite data sets 
treat_cascade_country_list <- list(treat_cascade_all, treat_cascade_f_adult, treat_cascade_m_adult)
names(treat_cascade_country_list) <- c("treat_cascade_all", "treat_cascade_f_adult", "treat_cascade_m_adult")

treat_cascade_global_list <- list(treat_cascade_all_global, treat_cascade_f_adult_global, treat_cascade_m_adult_global)
names(treat_cascade_global_list) <- c("treat_cascade_all_global", "treat_cascade_f_adult_global", "treat_cascade_m_adult_global")

test_treat_cascade_country_list <- list(test_treat_cascade_all, test_treat_cascade_f_adult, test_treat_cascade_m_adult)
names(test_treat_cascade_country_list) <- c("test_treat_cascade_all", "test_treat_cascade_f_adult", "test_treat_cascade_m_adult")

test_treat_cascade_global_list <- list(test_treat_cascade_all_global, test_treat_cascade_f_adult_global, test_treat_cascade_m_adult_global)
names(test_treat_cascade_global_list) <- c("test_treat_cascade_all_global", "test_treat_cascade_f_adult_global", "test_treat_cascade_m_adult_global")

# trim composite datasets
treat_cascade_country_list_trimmed <- map(treat_cascade_country_list, trim_treat_country)

treat_cascade_global_list_trimmed <- map(treat_cascade_global_list, trim_treat_global)

test_treat_cascade_country_list_trimmed <- map(test_treat_cascade_country_list, trim_test_treat_country)

test_treat_cascade_global_list_trimmed <- map(test_treat_cascade_global_list, trim_test_treat_global)

# separate into individual data frames
list2env(treat_cascade_country_list_trimmed, envir = .GlobalEnv)
list2env(treat_cascade_global_list_trimmed, envir = .GlobalEnv)
list2env(test_treat_cascade_country_list_trimmed, envir = .GlobalEnv)
list2env(test_treat_cascade_global_list_trimmed, envir = .GlobalEnv)

# add sex and age columns
list_all_ages <- list(treat_cascade_all, treat_cascade_all_global, test_treat_cascade_all, test_treat_cascade_all_global)
names(list_all_ages) <- c("treat_cascade_all", "treat_cascade_all_global", "test_treat_cascade_all", "test_treat_cascade_all_global")
map(list_all_ages, mutate, sex = "all", age = "all") %>%
  list2env(envir = .GlobalEnv)

list_m_adult <- list(treat_cascade_m_adult, treat_cascade_m_adult_global, test_treat_cascade_m_adult, test_treat_cascade_m_adult_global)
names(list_m_adult) <- c("treat_cascade_m_adult", "treat_cascade_m_adult_global", "test_treat_cascade_m_adult", "test_treat_cascade_m_adult_global")
map(list_m_adult, mutate, sex = "male", age = "adult") %>%
  list2env(envir = .GlobalEnv)

list_f_adult <- list(treat_cascade_f_adult, treat_cascade_f_adult_global, test_treat_cascade_f_adult, test_treat_cascade_f_adult_global)
names(list_f_adult) <- c("treat_cascade_f_adult", "treat_cascade_f_adult_global", "test_treat_cascade_f_adult", "test_treat_cascade_f_adult_global")
map(list_f_adult, mutate, sex = "female", age = "adult") %>%
  list2env(envir = .GlobalEnv)

# join datasets to create full cascade datasets for global datasets
full_cascade_all_global <- full_join(treat_cascade_all_global, 
                                     test_treat_cascade_all_global, 
                                     by = c("Region", "age", "sex"))

full_cascade_m_adult_global <- full_join(treat_cascade_m_adult_global,
                                         test_treat_cascade_m_adult_global, 
                                         by = c("Region", "age", "sex"))

full_cascade_f_adult_global <- full_join(treat_cascade_f_adult_global, 
                                         test_treat_cascade_f_adult_global, 
                                         by = c("Region", "age", "sex"))

# union global datasets to include sex and age data in a single dataset
full_cascade_global <- union(full_cascade_all_global,
                                 full_cascade_m_adult_global) %>%
                       union(full_cascade_f_adult_global)
```


```{r reshape and tidy global composite data}
# create function to tidy composite global dataset
tidy_composite <- function(data) {
  data %>%
    # replace "..." values with NA
    na_if("...") %>%
    # gather by metric types
    gather(key = "year_hiv_vl_suppress", value = "estimate_hiv_vl_suppress", ends_with(("_hiv_vl_suppress"))) %>%
    gather(key = "year_hiv_on_art", value = "estimate_hiv_on_art", ends_with(("_hiv_on_art"))) %>% 
    gather(key = "year_on_art_vl_suppress", value = "estimate_on_art_vl_suppress", ends_with(("_on_art_vl_suppress"))) %>%
    gather(key = "year_know_status_on_art", value = "estimate_know_status_on_art", ends_with(("_know_status_on_art"))) %>% 
    gather(key = "year_know_status", value = "estimate_know_status", ends_with(("_know_status"))) %>% 
    # remove "_on_art_vl_suppress", "_know_status_on_art", and "_know_status" from year columns
    map(str_replace_all, "_hiv_vl_suppress", "") %>% 
    map(str_replace_all, "_hiv_on_art", "") %>%
    map(str_replace_all, "_on_art_vl_suppress", "") %>% 
    map(str_replace_all, "_know_status_on_art", "") %>%
    map(str_replace_all, "_know_status", "") %>%
    # remove any "<" or ">" characters (just set to value)
    map(str_replace_all, "<", "") %>%
    map(str_replace_all, ">", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for all metrics
    filter(year_hiv_vl_suppress ==  year_hiv_on_art & 
           year_hiv_vl_suppress ==  year_on_art_vl_suppress &
           year_hiv_vl_suppress == year_know_status_on_art &
           year_hiv_vl_suppress == year_know_status) %>% 
    # gather all estimates and create a column for the metric type
    gather(key = "metric_type", value = "metric_value", starts_with("estimate_")) %>%
    # drop redundant year columns
    select(-year_hiv_vl_suppress, -year_hiv_on_art, -year_on_art_vl_suppress, -year_know_status_on_art) %>%
    # spread into estimate, upper, and lower columns
    spread(key = year_know_status, metric_value) %>%
    # gather all lower_ci estimates, upper_ci estimates, and center_estimates
    gather(key = "year_lower", value = "lower_ci", contains(("_lower"))) %>% 
    gather(key = "year_upper", value = "upper_ci", contains(("_upper"))) %>% 
    gather(key = "year", value = "estimate", starts_with("20")) %>% 
    # remove "_lower" and "_upper" suffixes to get years
    map(str_replace_all, "_lower", "") %>% 
    map(str_replace_all, "_upper", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for estimate, lower_ci, and "upper_ci"
    filter(year == year_upper & year == year_lower) %>% 
    # select and reorder columns
    select(1, year, metric_type, estimate, lower_ci, upper_ci, sex, age) %>%
    # convert columns to appropriate data classes
    mutate(year = as.factor(year),
           estimate = as.numeric(estimate),
           lower_ci = as.numeric(lower_ci),
           upper_ci = as.numeric(upper_ci))
}

# tidy global composite data sets
full_cascade_global_tidy <- tidy_composite(full_cascade_global)
write_rds(full_cascade_global_tidy, "data/full_cascade_global_tidy")
```

```{r reshape and tidy country composite data}
# create function to tidy country treat datasets
tidy_treat <- function(data) {
  data %>%
    # replace "..." values with NA
    na_if("...") %>%
    # gather by metric types
    gather(key = "year_on_art_vl_suppress", value = "estimate_on_art_vl_suppress", ends_with(("_on_art_vl_suppress"))) %>%
    gather(key = "year_know_status_on_art", value = "estimate_know_status_on_art", ends_with(("_know_status_on_art"))) %>% 
    gather(key = "year_know_status", value = "estimate_know_status", ends_with(("_know_status"))) %>% 
    # remove "_on_art_vl_suppress", "_know_status_on_art", and "_know_status" from year columns
    map(str_replace_all, "_on_art_vl_suppress", "") %>% 
    map(str_replace_all, "_know_status_on_art", "") %>%
    map(str_replace_all, "_know_status", "") %>%
    # remove any "<" or ">" characters (just set to value)
    map(str_replace_all, "<", "") %>%
    map(str_replace_all, ">", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for all metrics
    filter(year_on_art_vl_suppress == year_know_status_on_art &
           year_on_art_vl_suppress == year_know_status) %>% 
    # gather all estimates and create a column for the metric type
    gather(key = "metric_type", value = "metric_value", starts_with("estimate_")) %>%
    # drop redundant year columns
    select(-year_on_art_vl_suppress, -year_know_status_on_art) %>%
    # spread into estimate, upper, and lower columns
    spread(key = year_know_status, metric_value) %>%
    # gather all lower_ci estimates, upper_ci estimates, and center_estimates
    gather(key = "year_lower", value = "lower_ci", contains(("_lower"))) %>% 
    gather(key = "year_upper", value = "upper_ci", contains(("_upper"))) %>% 
    gather(key = "year", value = "estimate", starts_with("20")) %>% 
    # remove "_lower" and "_upper" suffixes to get years
    map(str_replace_all, "_lower", "") %>% 
    map(str_replace_all, "_upper", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for estimate, lower_ci, and "upper_ci"
    filter(year == year_upper & year == year_lower) %>% 
    # select and reorder columns
    select(1, year, metric_type, estimate, lower_ci, upper_ci, sex, age) %>%
    # convert columns to appropriate data classes
    mutate(year = as.factor(year),
           estimate = as.numeric(estimate),
           lower_ci = as.numeric(lower_ci),
           upper_ci = as.numeric(upper_ci))
}

# apply tidy_treat function to all country treatment cascade datasets
treat_cascade_country_list_tidy <- map(list(treat_cascade_all, treat_cascade_f_adult, treat_cascade_m_adult), tidy_treat)

# separate into individual data frames
names(treat_cascade_country_list_tidy) <- map_chr(names(treat_cascade_country_list), paste0, "_tidy")
list2env(treat_cascade_country_list_tidy, envir = .GlobalEnv)

# create function to tidy country test_treat datasets
tidy_test_treat <- function(data) {
  data %>%
    # replace "..." values with NA
    na_if("...") %>%
    # gather by metric types
    gather(key = "year_hiv_vl_suppress", value = "estimate_hiv_vl_suppress", ends_with(("_hiv_vl_suppress"))) %>%
    gather(key = "year_hiv_on_art", value = "estimate_hiv_on_art", ends_with(("_hiv_on_art"))) %>% 
    # remove "_on_art_vl_suppress", "_know_status_on_art", and "_know_status" from year columns
    map(str_replace_all, "_hiv_vl_suppress", "") %>% 
    map(str_replace_all, "_hiv_on_art", "") %>%
    # remove any "<" or ">" characters (just set to value)
    map(str_replace_all, "<", "") %>%
    map(str_replace_all, ">", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for all metrics
    filter(year_hiv_vl_suppress ==  year_hiv_on_art) %>% 
    # gather all estimates and create a column for the metric type
    gather(key = "metric_type", value = "metric_value", starts_with("estimate_")) %>%
    # drop redundant year columns
    select(-year_hiv_vl_suppress) %>%
    # spread into estimate, upper, and lower columns
    spread(key = year_hiv_on_art, metric_value) %>%
    # gather all lower_ci estimates, upper_ci estimates, and center_estimates
    gather(key = "year_lower", value = "lower_ci", contains(("_lower"))) %>% 
    gather(key = "year_upper", value = "upper_ci", contains(("_upper"))) %>% 
    gather(key = "year", value = "estimate", starts_with("20")) %>% 
    # remove "_lower" and "_upper" suffixes to get years
    map(str_replace_all, "_lower", "") %>% 
    map(str_replace_all, "_upper", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for estimate, lower_ci, and "upper_ci"
    filter(year == year_upper & year == year_lower) %>% 
    # select and reorder columns
    select(1, year, metric_type, estimate, lower_ci, upper_ci, sex, age) %>%
    # convert columns to appropriate data classes
    mutate(year = as.factor(year),
           estimate = as.numeric(estimate),
           lower_ci = as.numeric(lower_ci),
           upper_ci = as.numeric(upper_ci))
}

# apply tidy_test_treat function to all country treatment cascade datasets
test_treat_cascade_country_list_tidy <- map(list(test_treat_cascade_all, test_treat_cascade_f_adult, test_treat_cascade_m_adult), tidy_test_treat)

# separate into individual data frames
names(test_treat_cascade_country_list_tidy) <- map_chr(names(treat_cascade_country_list), paste0, "_tidy")
list2env(test_treat_cascade_country_list_tidy, envir = .GlobalEnv)

# create union of datasets to create full cascade for country datasets
full_cascade_all_tidy <- union(treat_cascade_all_tidy, 
                          test_treat_cascade_all_tidy)

full_cascade_m_adult_tidy <- union(treat_cascade_m_adult_tidy, 
                              test_treat_cascade_m_adult_tidy)

full_cascade_f_adult_tidy <- union(treat_cascade_f_adult_tidy, 
                              test_treat_cascade_f_adult_tidy)

# union country datasets to include sex and age data in a single dataset
full_cascade_country_tidy <- union(full_cascade_all_tidy,
                                   full_cascade_m_adult_tidy) %>%
                             union(full_cascade_f_adult_tidy)

# save RDS object
write_rds(full_cascade_country_tidy, "data/full_cascade_country_tidy")
```


```{r reshape and tidy raw single metric data}
# create function to tidy single metric datasets
tidy_single <- function(data) {
  data %>%
    # replace "..." values with NA
    na_if("...") %>%
    # gather all lower_ci estimates, upper_ci estimates, and center_estimates
    gather(key = "year_lower", value = "lower_ci", contains(("_lower"))) %>% 
    gather(key = "year_upper", value = "upper_ci", contains(("_upper"))) %>% 
    gather(key = "year", value = "estimate", starts_with("20")) %>% 
    # remove "any suffixes "_upper" and "_lower"
    map(str_replace_all, "_lower", "") %>% 
    map(str_replace_all, "_upper", "") %>%
    # remove any "<" or ">" characters (just set to value)
    map(str_replace_all, "<", "") %>%
    map(str_replace_all, ">", "") %>%
    # convert vectorized data into a tibble
    as_tibble() %>% 
    # filter so that years match for estimate, lower_ci, and "upper_ci"
    filter(year == year_upper & year == year_lower) %>% 
    # select and reorder columns
    select(1, year, estimate, lower_ci, upper_ci) %>%
    # convert columns to appropriate data classes
    mutate(year = as.factor(year),
           estimate = as.numeric(estimate),
           lower_ci = as.numeric(lower_ci),
           upper_ci = as.numeric(upper_ci))
}

      
# create list of single metric datasets
single_data_list <- list(know_status_all_global, on_art_all_global, vl_suppress_all_global, know_status_all, on_art_all, vl_suppress_all)

single_list_names <- c("know_status_all_global", "on_art_all_global", "vl_suppress_all_global", "know_status_all", "on_art_all", "vl_suppress_all")

names(single_data_list) <- single_list_names

# tidy single metric data sets
single_data_list_tidy <- map(single_data_list, tidy_single)
names(single_data_list_tidy) <- map_chr(single_list_names, paste0, "_tidy")

# split lists into individual tidy data frames
list2env(single_data_list_tidy, envir = .GlobalEnv)
```


```{r}
# plot data sets
know_status_all_global_tidy %>% 
  ggplot(aes(x = year, y = estimate, color = Region, group = Region)) + 
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept = 90, alpha = 0.2) + 
  scale_y_continuous(labels = percent_format(scale = 1), 
                     breaks = pretty_breaks(n = 5), 
                     expand = expansion(.2)) +
  theme(panel.grid = element_blank(),
        panel.background = element_blank()) +
  scale_color_viridis_d()
```

To Do:
-rename metric_type to measure in functions
-drop "estimate_" prefix from measure column
-add some rm() functions to clean stray objects from the environment
-check if single metric sets match filtered versions of composite datasets
-create single metric datasets for missing measures by subsetting composite datasets

